A multi-scale story of the diffusion of a new technology: the web

### 2025-01-21 paper/submission2/tranos2025.qmd

<!-- geographers lack of interest for diffusion + motivation -->
Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of 
@hagerstrand1968innovation, which demonstrated the importance of communications 
as a mechanism for the diffusion of innovations, being the most impactful. 
It was followed by various empirical studies -- see for instance @ormrod1990 
and @iso2005. However, the torch of exploring and modelling such diffusion 
processes had been passed to other disciplines such as economics, business 
studies and sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more specifically for 
this paper, digital technologies across both space and time can be attributed to 
the scarcity of relevant and granular enough data, a problem also highlighted 
by @iso2005 and @kemeny2011international. As @zook2022mapping pointed, digital 
activities are hardly ever captured in official data.

<!-- contribution -->
This paper offers such a contribution: an economic geography study revealing
the importance of spatial mechanisms for the diffusion of an intangible, 
digital technology that is the Web in the UK at a high level of spatial 
granularity during the 1996-2012 period.
<!-- how a new technology that is the Web diffused over space and time  -->
It does so by employing a novel source of big data and machine 
learning algorithms to model the *active* engagement with Web during that 
period and tests how different spatial and temporal mechanisms shaped its 
diffusion. This paper also exemplifies how the combination of data sources 
and state-of-the-art research methods which escape the traditional social 
science domain can offer new lenses to geographical research regarding the 
understanding of technological diffusion.

@@

<!-- motivation: regional development -->
This gap becomes even more prominent if we consider the importance of 
technology and its adoption in economic geography and macroeconomic growth 
theories. Simply put, differences in technological adoption and sophistication 
may lead to uneven development as new technologies can increase productivity 
and allow firms to introduce new products [@solow1957technical; 
@aghion1990model; @kemeny2011international]. Linked to the subject of this 
paper, the Web -- and the active engagement with this technology that is 
creating and maintaining instead of just browsing a website -- can be 
considered as the underpinning technology for the transformative process 
identified by @capello2024nexus as the *digital service economy*. Importantly, 
the focus here is not the Web in general, but instead commercial websites.
After all, diffusion together with invention and innovation are considered the 
pillars of technological change [@das2022diffusion]. Findings from previous 
studies illustrated that early engagement with digital technologies such as 
the Web led to longstanding positive productivity gains for these regions that 
hosted early adopters [@tranosuk]. Hence, obtaining a robust understanding of 
the spatial diffusion mechanisms is crucial in mapping future development 
trajectories.

<!-- motivation: intangible/digital -->
Another line of motivation has to do with the intangible nature of the Web 
and the debate about the footloose nature of digital technologies, which
contrary to several earlier general purpose technologies are 'light in weight' 
[@coyle1999weightless]. Following the seminal work of @gaspar1998information, 
which demonstrated that technological improvements in telecommunications lead 
to increased demand for face-to-face communications and, consequently, 
enhance the importance of cities, empirical studies from economic geography 
and urban economics explored whether digital technologies substitute or 
complement agglomeration externalities and spatial structure 
[@kolko_death_2000; @ioannides2008effect; @tranos2021ubiquitous].
With the exception of @sinai2004geography, who focused, among other things, 
on web content and demonstrated that more such content is available in larger
markets, this strand of literature has been mostly concerned with the
spatiality of the Internet's hard and tangible infrastructure
and shied away from exploring the spatial dimensions of one of the 
most distinct software layers of the Internet infrastructure, the Web.

<!-- diffusion and adoption -->
Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, @rogers2010diffusion identifies early adopter of new
technologies as 'knowledgeable risk takers' and @griliches1957 as
'profit maximisers' [@ding2010modeling]. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, @perkins2011internet attribute diffusion
to two processes: (i) epidemic-like mechanisms, which are governed by
distance, proximity and social interactions, and (ii) economic
mechanisms as new innovations are adopted by users as they become more
profitable, valueable and useful.
<!-- The criticisms are mainly based on the fact that although the approach -->
<!-- gives an idea of aggregate (industry or household) behaviour, it does -->
<!-- not focus on the individual's (firm or household) adoption process -->

@@

<!-- data -->
The data used here depict the active engagement with the Web as they contain 
geolocated and time-stamped counts of commercial websites from the UK. Instead 
of adopting metrics reflecting the passive engagement with the Internet such as 
Internet adoption and speeds [e.g. @blank2018local; @destefano2022fuzzy], this 
paper observes yearly counts of geolocated websites, which reflect the creation 
and maintenance of websites. It needs to be highlighted here that although 
there are various types of underpinning web technologies -- namely markup, 
stylesheet, client- and server-side coding languages as well as different
web browsers and other related programming language -- the focus here
is on the diffusion of the Web in its totality. For a more detailed and 
cross-country technological diffusion study, see @PAPAGIANNIDIS2015308.

<!-- description of the methods and results -->
To analyse these data, a diverse set of methods are employed. Firstly, 
cumulative adoption curves are estimated for the UK as a whole and for the 
Local Authority Districts (LAD), illustrating, for the first time, such 
S-shaped diffusion patterns for local areas. This system level analysis also 
reveals insights about the volatility of the adoption of the Web. Then, 
exploratory spatial data analysis offers some first indications of the 
importance of the different spatial diffusion mechanisms in shaping the 
diffusion of the Web in the UK. Importantly, the analysis adopts two distinct 
geographical scales: the LAD and the most detailed scale of UK census areas, 
the Output Areas (OA). Lastly, a novel methodological framework based on 
Machine Learning (ML) algorithms is developed to test the 
role of such mechanisms at these two scales. The results reveal, for the first 
time, the importance of different spatial diffusion mechanisms in the diffusion 
of an intangible, digital technology at very granular geographical scales.

<!-- contents -->
The structure of the paper goes as follows. @sec-litreview reviews the key
concepts in the diffusion literature as well as relevant empirical studies. 
Then, @sec-datamethods describes the data and methods used in this paper. 
The results of the different stages of the analysis are presented in 
@sec-results. The discussion of the results and the conclusions of the paper
are offered in @sec-conclusions.

<!-- @grubler1990rise Later Hagerstrand conceptualized physical "barrier" -->
<!-- effects like lakes or uninhabited areas, which, in addition to distance, -->
<!-- act as further retarding effects on diffusion. These are formalized in -->
<!-- the form of "zero" or "half" contact multiplicators on the (distance -->
<!-- decaying) message flows. -->

<!-- @grubler1990rise With respect to the formalization of the communication -->
<!-- flows Hagerstrand defines a "mean information field" (MIF), in which the -->
<!-- probability of communication is a negative function of distance between -->
<!-- individuals -->

# Literature review {#sec-litreview}

<!-- key concepts -- copied from help.Rmd -->
Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be 
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovations just like pathogens spread because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, there is a hierarchical processes. According to this, instead of 
a horizontal distance-based diffusion mechanism, the top-down hierarchy
of urban systems shapes technological diffusion. In reality, the synthesis 
of these two processes better reflects how new technologies diffuse over 
space [@morrill2020spatial].

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled 'Innovation Diffusion as a Spatial Process'
[@hagerstrand1968innovation]. Hägerstrand was the first to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified
a 'neighbourhood' mechanism of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the 'lead-lag'
spatial acceleration effect in market research [@bento2018time;
@PERES201091]. Hägerstrand is more widely known though for highlighting
the role time plays in the diffusion of innovations: an early-pioneering
period, a middle fast accelerating period and a final saturation one
[@morrill2020spatial].

The temporal dimension was further explored by Everett Rogers and his
seminal work on 'Diffusion of Innovations' [@rogers2010diffusion].
Rogers being a sociologist focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism of
diffusion and adoption is communication and how knowledge is transferred 
within a social system. All above approaches agree that diffusion and the
cumulative adoption of a new technology are represented by an S-shaped 
line plot [@grubler1990rise].

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing].

@grubler1990rise effectively summarises the three key spatial diffusion 
mechanisms: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchical mechanism in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and proximity. 
These are the three mechanisms that the empirical analysis in @sec-results 
investigates.

<!-- previous studies -->
The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metro areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across 402 German regions during 1970-2015. Their
key finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a
region or industry with high share of firms having a website, they are
more likely to have a website too.

At a global scale @perkins2005international explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
@perkins2011internet explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the Internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, @ding2010modeling modelled the spatial diffusion of
mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network, @lengyel2020role
analysed its adoption and the churn at a very granular spatial level.
Their results were in agreement with the early theoretical and empirical
contributions reviewed here: assortativity, urban scaling and distance
are the key drivers of spatial diffusion. @bakher2013diffusion compared 
different the growth curves as a means to model diffusion of online 
shopping in Australia during the 1998-2009 period. At a global scale
@PAPAGIANNIDIS2015308 modelled the diffusion of different web
technologies and practices across countries. Using similar but less extensive 
data as the one used here, their analysis illustrated how the diffusion 
of different web technologies and practices follow an S-shaped pattern 
as well as the different diffusion rates of the different technologies 
and practices.

All in all, this paper tests whether the three distinct spatial diffusion 
mechanisms shaped the diffusion of the Web in the UK at very granular
spatial scales from the Web's early stages onwards. The next section discusses
the empirical framework.

<!-- @fritsch2015new analyze the spatial diffusion of laser technology -->
<!-- research in West Germany from 1960, when this technology began, until -->
<!-- 2005. -->
<!-- in one region  -->
<!-- @leibowicz2016representing for energy There is no generally accepted -->
<!-- theory that explains diffusion rate heterogeneity across technologies, -->
<!-- but several factors are considered important. Greater unit scale and -->
<!-- larger market size contribute to slower diffusion. Requirements for -->
<!-- interrelated technologies or complex infrastructures also hinder the -->
<!-- diffusion process (Grubler, 2012). -->

<!-- Mobile phones benefited from early deployment in recreational boats and -->
<!-- automobiles, where the traditional competitor was not a viable option. -->
<!-- In the early stages of diffusion, performance is a more important driver -->
<!-- of adoption than cost competitiveness. Typically, significant cost -->
<!-- reductions only occur once the technology reaches a deployment level -->
<!-- capable of supporting standardization and mass production (Wilson, -->
<!-- 2012). -->

<!-- @leibowicz2016representing Empirical evidence supports the validity of -->
<!-- Schmidt's Law over a wide range of technologies, time periods, and -->
<!-- geographical contexts. A recent meta-analysis of technology up-scaling -->
<!-- found that diffusion accelerated moving from the core to the rim and -->
<!-- periphery for technologies as diverse as natural gas power, oil -->
<!-- refineries, and automobiles (Wilson, 2009). One historical example that -->
<!-- conforms particularly well to Schmidt's Law is the diffusion of coal -->
<!-- power in Europe (Grubler, 2012). England emerged as the core region for -->
<!-- coal power because it had legal and economic institutions that -->
<!-- incentivized scientific pursuits, domestic coal reserves, and a clear -->

<!-- industrial motivation to replace water power with coal. ... -->

<!-- country level, not sure if needed. -->

<!-- @bento2018time explore What determines the duration of formative phases -->
<!-- for energy innovations in different markets? We are interested both in -->
<!-- initial markets (also: core, lead, first mover, early adopter) where -->
<!-- formative phases prepare technologies for mass commercialization, and in -->
<!-- follower markets (also: periphery, lag, late adopter) where accelerated -->
<!-- formative phases may benefit from diffusion and spillovers. -->

@@

<!-- main metric -->
To capture the diffusion of the Web, a website density metric is developed 
for two different geographical scales: the Local Authority Districts (LAD) 
and the Output Areas (OA). The former is an administrative unit and there 
are 374 LAD in the UK. The latter is the smallest census-based geographical 
unit and there are c. 230,000 of them in the UK. This methodological choice 
will allow the mapping of the diffusion of web technologies and the assessment 
of the diffusion mechanisms at these two very different spatial scales.

<!-- web data -->
The counts of websites at these scales are calculated using data from the 
Internet Archive[^2] and, specifically, the JISC UK Web Domain Dataset 
[@ukwebarchive]. The Internet Archive is one of the most complete and oldest 
archive of webpages in the world operating since 1996 
[@ainsworth2011much; @holzmann2016dawn]. It is a web crawler, which discovers 
webpages by following the hyperlinks of every webpage it archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK country code top level domain (ccTLD - .uk) 
from the 1996–2012 period. In essence, this is a long list of 2.5 billion URLs 
of archived webpages including also the archival timestamp.

@@

Instead of using the whole .uk ccTLD, this paper focuses on its commercial 
subset, the .co.uk second level domain (SLD). This choices decrease the 
heterogeneity of the web data as such commercial websites have specific 
aims: they are used to diffuse information, support online transactions 
and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as .com and these 
cases escape the data used here, such omissions should not affect the 
validity of the results given the popularity of the .uk ccTLD [@tranosuk]: 
UK consumers prefer to visit a .uk website when they are searching for 
products or services [@hope]; and anecdotal evidence indicates that during 
the first half of 2000, three .co.uk domains were registered every minute 
[@oecd_coms]. Importantly, previous studies illustrated that .co.uk is the 
most popular UK SLD [@tranosuk].

The text from these webpages was scanned using a regular expression (regex) to
identify strings of text which resemble UK postcodes and one fifth of them 
included a mention to a postcode [@BL2013geo]. This information allows the 
geolocation of the data and the creation of the LAD and OA counts.

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This website 
reconstruction process lend itself to the construction of *website* instead 
of *webpage* density metrics. Websites tend to represent specific organisations 
or entities, and, arguably, are more meaningful observational units than 
webpages, which can ignore the upstream dependency of the website they belong 
to. Based on the following example, all three webpages are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)). Only 
the overall website and not the nested webpages are considered for the counts 
as otherwise the density metrics would have been biased towards large, place 
specific, websites.

  - [http://www.website.co.uk/webpage_a](http://www.website.co.uk/webpage_a) B15 2TT

  - [http://www.website.co.uk/webpage_b](http://www.website.co.uk/webpage_b) BS8 1TH

  - [http://www.website.co.uk/webpage_c](http://www.website.co.uk/webpage_c) B15 2TT

\noindent What is challenging is that this aggregation approach,
which has been used elsewhere [@tranosuk; @shoreditch] may lead to websites with 
multiple postcodes. As per the above example, 
[www.website.co.uk](http://www.website.co.uk) includes two unique postcodes: 
B15 2TT and BS8 1SS. The distribution of the number of postcodes per website for 
2000 is presented in Table \ref{f2000}, which illustrates a wide range. 
At the left end of the distribution, there are websites anchored to a unique 
location (72% of all the reconstructed websites in 2000), which may represent a 
small company with a single trading location. At the right end, there are 
websites with thousands of different postcodes. Considering the time period of 
the analysis, such cases can represent directories which used to be popular 
in the pre-search engines early times of the commercial Internet as well as real 
estate websites [@tranosuk].

@@

To create the website density metric, the yearly website counts at the LAD level 
are standardised by the number of firms in LAD to avoid biases associated with 
LAD hosting a large number of firms. Given that there is no such readily 
available statistic for the OA, the actual OA level counts are used. However, 
because of the the consistent spatial definition of OA (they host 40-250 
households),[^4] website counts in OA are interpreted as a density metric too.

[^4]: [https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies](https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies)

>>>>

<!-- ADD SOMETHING ABOUT MULTIPLE SCALES -->

<!-- geographers lack of interest for diffusion + motivation -->
Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
@hagerstrand1968innovation, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance @ormrod1990 and @iso2005. However, the torch of exploring
and modelling such diffusion processes had been passed to other
disciplines such as economics, business studies and sociology well
before the 'cultural turn' of economic geography
[@perkins2005international; @ding2010modeling].
<!-- NEW spatial granularity --> As a consequence, relevant studies
did not investigate diffusion mechanisms at granular geographical scales to 
provide local policy insights, for instance, regarding infrastructural d
evelopment and local technology policies. 
A potential explanation of the lack of geographical studies exploring the 
diffusion of new and, more specifically for this paper, digital technologies 
across both space and time can be attributed to the scarcity of relevant and 
granular enough data, a problem also highlighted by @iso2005 and
@kemeny2011international. As @zook2022mapping pointed, digital
activities are hardly ever captured in official data.

<!-- contribution -->
This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of an
intangible, digital technology that is the Web in the UK at a high level
of spatial granularity 
<!-- NEW TO JUSTIFY DATA TIME PERIOD-->
from the very early days until its maturity (1996-2012).
<!-- NEW DATA CLARIFICATION -->
It specifically focuses on the diffusion of commercial websites.
It does so by employing a novel source of big data and machine learning
algorithms to model the *active* engagement with Web -- that is creating and 
maintaining instead of just browsing a website -- during that period
and tests how different spatial and temporal mechanisms shaped its
diffusion. This paper also exemplifies how the combination of data
sources and state-of-the-art research methods which escape the
traditional social science domain can offer new lenses to geographical
research regarding the understanding of technological diffusion.

@@

<!-- motivation: regional development -->
This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
[@solow1957technical; @aghion1990model; @kemeny2011international].
Linked to the subject of this paper, the Web and, more specifically, 
commercial websites can be considered as the underpinning
technology for the transformative process identified by
@capello2024nexus as the *digital service economy*. Importantly, the
focus here is not the Web in general, but instead commercial websites.
After all, diffusion together with invention and innovation are
considered the pillars of technological change [@das2022diffusion].
Findings from previous studies illustrated that early engagement with
digital technologies such as the Web led to longstanding positive
productivity gains for these regions that hosted early adopters
[@tranosuk]. Hence, obtaining a robust understanding of the spatial
diffusion mechanisms is crucial in mapping future development trajectories.

<!-- NEW motivation: intangible/digital -->
Another line of motivation has to do with the very nature of the Web as a 
digital technology and how it evolved over time. The digital geographies field 
has debunked premature claims about the Internet's placeless-ness by mapping 
various of its facets and technological layers [@tranos2013death], but the lack 
of relevant long time-series data prevented researchers from observing digital
phenomena overtime. Most of these studies focused on its hard infrastructure 
[e.g. @malecki2002economic; @moss2000internet], on Internet subscribers 
[e.g. @blank2018local], or on social media users [e.g. @crampton2013beyond].
Exceptions include early work from @zook2000web, @zook2000web and 
@moss1997tracking, which utilised domain name registration data and use it to
map and illustrate the spatial footprint of the web content. Pivotal as these 
contributions have been in understanding the economic geographies of the 
Internet, they were based on cross-sectional, often spatially detailed, mapping 
and on shallow time comparisons (e.g. 1993-1998) and did not observe the 
evolution and diffusion of the Web until its maturity. As we know from the 
relevant literature, different technologies diffuse at different rates and 
there is no theory that can explain these differences 
[@leibowicz2016representing]. Hence, mapping and modelling the diffusion of the
Web among firms can offer useful insights.

@@

<!-- data -->
The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the Internet such as Internet adoption and speeds [e.g.
@blank2018local; @destefano2022fuzzy], this paper observes yearly counts
of geolocated, commercial websites, which reflect the creation and 
maintenance of such websites. It needs to be highlighted here that although 
there are various types of underpinning web technologies -- namely markup,
stylesheet, client- and server-side coding languages as well as
different web browsers and other related programming language -- the
focus here is on the diffusion of the Web as an overarching technology
instead of more specific web technologies. For a such a detailed and 
cross-country technological diffusion study, see @PAPAGIANNIDIS2015308.

<!-- description of the methods and results -->
<!-- AMMEND I AFTER CHANGE THE RESULTS -->
To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. This system level
analysis also reveals insights about the volatility of the adoption of
the Web. Then, exploratory spatial data analysis offers some first
indications of the importance of the different spatial diffusion
mechanisms in shaping the diffusion of the Web in the UK. Importantly,
the analysis adopts two distinct geographical scales: the LAD and the
most detailed scale of UK census areas, the Output Areas (OA). Lastly, a
novel methodological framework based on Machine Learning (ML) algorithms
is developed to test the role of such mechanisms at these two scales.
The results reveal, for the first time, the importance of different
spatial diffusion mechanisms in the diffusion of an intangible, digital
technology at very granular geographical scales.

<!-- contents -->
<!-- AMMEND I AFTER CHANGE THE RESULTS -->
The structure of the paper goes as follows. @sec-litreview reviews the
key concepts in the diffusion literature as well as relevant empirical
studies. Then, @sec-datamethods describes the data and methods used in
this paper. The results of the different stages of the analysis are
presented in @sec-results. The discussion of the results and the
conclusions of the paper are offered in @sec-conclusions.

# The mechanisms that shape the diffusion of a new technology {#sec-litreview}

<!-- The fundamentals of technology diffusion -->
The diffusion of new technologies and innovations is driven by three
different mechanisms [@grubler1990rise; @morrill2020spatial]: (i)
distance and proximity, (ii) urban hierarchies in the form of a
centrifugal forces as new technologies diffuse from core to periphery,
and (iii) over time the cumulative level of adoption follows an S-shaped
pattern.
<!-- neighbourhood -->
@hagerstrand1968innovation was the first to identify diffusion as a
geographical process. The starting point was the idea that diffusion is
based on passing information through social networks, which themselves
tend to be geographically bounded. Hence, he identified a
'neighbourhood' mechanism of how information, and consequently,
innovations and new technologies diffuse. This perspective draws
similarities with epidemics: new technologies and innovations, just like
pathogens, spread because of contagion and, consequently, proximity and
exposure [@hivner2003facilitating].

<!-- hierarchy -->
@hagerstrand1968innovation also emphasised the role of the top-down
hierarchy of urban systems and highlighted that new technologies and
innovations are firstly adopted in larger cities and then diffuse to
second tier ones. This is a sequential instead of a simultaneous
process, which resembles the 'lead-lag' spatial acceleration effect in
market research [@bento2018time; @PERES201091]. Schmidt's Law
empirically illustrates a similar pattern [@grubler1990rise]. Core and
highly agglomerated regions is where new technologies are invented and
commercially deployed. This is where the first adopters tend to be
based. Then, technologies spread to the rim and eventually to the
periphery. Although adoption pace might be higher when new technologies
finally arrive to the periphery, the saturation levels there may never
reach the ones in the core because of the lack of infrastructure or
other necessary institutions [@leibowicz2016representing].

<!-- time -->
Diffusion is naturally a dynamic process and Hägerstrand recognised the
role time plays in the diffusion of innovations and new technologies: an
early-pioneering period, a middle fast accelerating period and a final
saturation one [@morrill2020spatial]. This was further explored by
@rogers2010diffusion, who identified five types of adopters of new
technologies and innovations: innovators, early adopters, early
majority, late majority and laggards. Hence, plotting the cumulative
adoption of a new technology over time often leads to an S-shaped line
plot [@grubler1990rise].

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge gap
regarding the explanatory power of these mechanisms at granular, local
scales [@ding2010modeling]. As discussed in @sec-introduction, such 
spatially explicit knowledge can directly inform local technology
and infrastructure policies. Indeed, while the spatial dimension is
present in the following studies, the geographical level of analysis is
often too coarse to ascertain the role of the above mechanisms at the
local scale. @beardsell1999spatial studied the evolution of the computer
industry in 317 US metropolitan areas during the 1977-1992 period using
employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, @bednarz2020pulled focused on wind turbines and modelled their
spatial diffusion across 402 German regions during 1970-2015. Their key
finding is that local demand than local supply was the main driving
factor. @haller2011determinants employed firm panel data for Ireland
during 2003-2005 and illustrated that when firms are located in a region
or industry with high share of firms having a website, they are more
likely to have a website too. At a global scale @perkins2011internet
explored whether the adoption of previous communication technologies
that is mail, telegrams and telephones were shaped by similar
socioeconomic factors as the Internet. Their results indicated common
patterns regarding the drivers behind the adoption of different
communication technologies. @feldman2015rknowledge focused on inventions
and used patent data to map how the the rDNA technology diffused among
inventors across 366 US metropolitan areas. @ding2010modeling modelled
the spatial diffusion of mobile telecommunications across the highest
level of administrative division in China (29 regions in total). Their
analysis indicated that socioeconomic characteristics are important
determinants of the timing, speed and the level of mobile diffusion
within China. @bakher2013diffusion compared different the growth curves
as a means to model the diffusion of online shopping in Australia during
the 1998-2009 period, but they did not consider any spatial mechanisms.
At a global scale @PAPAGIANNIDIS2015308 modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
@lengyel2020role, who analysed the adoption and churn of a Hungarian
online social network. Their results were in agreement with the early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion.
Based on the above, the main empirical hypothesis states:

> H~1~: The three diffusion mechanisms namely distance, urban hierarchy
and the the S-shaped pattern of the cummulative level of adoption shaped
the diffusion of the commercial Web in the UK even at granular geographical 
scales.

<!-- EARLY/LATE adoption-->
Understanding the diffusion of a new technology in multiple, detailed spatial
scales is of interest to economic geography as potential local economic 
benefits may occur because of concentration of early adopters or through 
technological leapfrogging. On the one hand, early adopters are often rewarded 
because of their attitudes towards new technologies and innovations. 
@rogers2010diffusion identifies them as 'knowledgeable risk takers' and 
@griliches1957 as 'profit maximisers' [@ding2010modeling]. On the other hand,
latecomers can adopt and benefit from new technologies that have been developed
elsewhere without incurring the hefty initial R&D costs [@teece2008firm]. At a 
global scale @perkins2005international explored whether the diffusion rate of 
new technologies is driven by a latecomer advantage and the engagement with the 
global economy via foreign direct investments and trade. Their results 
illustrate that indeed latecomers and developing countries experience diffusion 
of new technologies more rapidly than early adopters and developed countries. 
At a regional scale, previous research highlighted the long term and 
sustained productivity benefit of the early adopters of web technologies in 
the UK [@tranosuk]. Although the early adoption and leapfrogging literature has 
not paid much attention on cities and regions [@yu2018sustainability], the 
potential local economic benefits justify the need to investigate the
dynamics of the web diffusion in different scales in the UK. Hence, the second 
empirical hypothesis states:

> H~2~: The diffusion of the commecrial Web is shaped by both early adoption 
and late comer effects.

These are the two hypotheses are empirically tested in @sec-results using the 
data and methods described in the next section.

<!-- Technological diffusion, which is by definition an aggregated process, -->
<!-- can be discussed in parallel with individual adoption mechanisms. On the -->
<!-- one hand,  -->
<!-- On the other hand, @perkins2011internet attribute diffusion -->
<!-- to two processes: (i) epidemic-like mechanisms, which are governed by -->
<!-- distance, proximity and social interactions, and (ii) economic -->
<!-- mechanisms as new innovations are adopted by users as they become more -->
<!-- profitable, valueable and useful. -->

@@

<!-- main data -->
To capture the diffusion of the commercial Web, a website density metric is
developed for two different geographical scales that are discussed later in 
this section. The main data are counts of commercial websites that are 
calculated using data from the Internet Archive[^2] and, specifically, the 
JISC UK Web Domain Dataset [@ukwebarchive]. The Internet Archive is one of the 
most complete and oldest archive of webpages in the world operating since
1996 [@ainsworth2011much; @holzmann2016dawn]. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996–2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

@@

Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choice
is aligned with the topic of the paper -- the diffusion of the Web for 
commercial functions -- and decreases the heterogeneity of the web data as such 
commercial websites have specific aims: they are used to diffuse information, 
support online transactions and share opinions [@THELWALL2000441; 
@blazquez2018big]. Although a UK company can adopt a generic TLD such as .com 
and these cases escape the data used here, such omissions should not affect the
validity of the results given the popularity of the .uk ccTLD
[@tranosuk]: UK consumers prefer to visit a .uk website when they are
searching for products or services [@hope]; and anecdotal evidence
indicates that during the first half of 2000, three .co.uk domains were
registered every minute [@oecd_coms]. Importantly, previous studies
illustrated that .co.uk is the most popular UK SLD [@tranosuk]. Henceforward 
any reference to websites will refer to commercial -- .co.uk -- websites.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode [@BL2013geo]. This
information allows the geolocation of the data and the creation of the
LAD and OA counts. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files [@ons].

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This
website reconstruction process lend itself to the construction of
*website* instead of *webpage* density metrics. Websites tend to
represent specific organisations or entities, and, arguably, are more
meaningful observational units than webpages, which can ignore the
upstream dependency of the website they belong to. Based on the
following example, all three webpages are part of the same website
(<http://www.website.co.uk>). Only the overall website and not the
nested webpages are considered for the counts as otherwise the density
metrics would have been biased towards large, place specific, websites.
It is important to note here that websites commonly contain place references, 
with previous research estimating this share to be around 70% 
[@hill2009georeferencing].

-   <http://www.website.co.uk/webpage_a> B15 2TT

-   <http://www.website.co.uk/webpage_b> BS8 1TH

-   <http://www.website.co.uk/webpage_c> B15 2TT

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere [@tranosuk; @shoreditch] may lead to websites
with multiple postcodes. As per the above example,
[www.website.co.uk](http://www.website.co.uk) includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of the number of
postcodes per website for 2000 is presented in Table \ref{f2000}, which
illustrates a wide range. At the left end of the distribution, there are
websites anchored to a unique location (72% of all the reconstructed
websites in 2000), which may represent a small company with a single
trading location. At the right end, there are websites with thousands of
different postcodes. Considering the time period of the analysis, such
cases can represent directories which used to be popular in the
pre-search engines early times of the commercial Internet as well as
real estate websites [@tranosuk]. Previous research which used the same 
methodology for a small subset of these data illustrates the robustness of the 
above approach: its results matched in terms of accuracy past rounds of 
research based on extensive qualitative work, interviews and observational 
studies for the same study area [@shoreditch].

@@

<!-- Multi-scale -->
To empirically test the two hypotheses stated in @sec-litreview, τhe website 
density metric was calculated for two different scales the Local Authority
Districts (LAD) and the Output Areas (OA). The former is an administrative unit 
and there are 374 LAD in the UK. The latter is the smallest census-based 
geographical unit and there are c. 230,000 of them in the UK. To create the 
website density metric, the yearly website counts at the LAD level are 
standardised by the number of firms in LAD to avoid biases associated with LAD 
hosting a large number of firms. Given that there is no such readily available 
statistic for the OA, the actual OA level counts are used. However, because of 
the the consistent spatial definition of OA (they host 40-250 households),[^3] 
website counts in OA are interpreted as a density metric too.

[^3]: <https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies>
